Alone Together: Compositional Reasoning and Inference for Weak Isolation

### 2017-11-08 inference.tex
Updated proof sketch in Section 5

We now formally state the correspondence between the inference rules
given above and the RG judgment of \S\ref{sec:reasoning}:
\begin{theorem}
  \label{thm:inference-sound}
  For all $i$,$R$,$I$,$c$,$\Fx$, $\F$, if $\stable(\R,I)$ and $\Fx
  \vdash c \elabsto \F$, then:\\\vspace*{-0.2cm}
  \begin{smathpar}
  \begin{array}{c}
    \R \vdash \hoare{\lambda(\stl,\stg).~\stl=\Fx(\stg) \conj
    I(\stg)}{c}{\lambda(\stl,\stg).~\stl = \Fx(\stg) \cup \F(\stg)
    \conj I(\stg)}
  \end{array}
  \end{smathpar}
\end{theorem}

\paragraph{{\sc Proof Sketch.}} 
% We prove a slightly stronger statement :
% For all  $i$,$R$,$I$,$c$,$\F$,$s$ if $\stable(\R,I)$ and $c \elabsto \F$, then:
%   \begin{smathpar}
%     \begin{array}{c}
%     \R \vdash \hoare{\lambda(\stl,\stg).~\stl=s \conj
%     I(\stg)}{c}{\lambda(\stl,\stg).\stl = \F(s,\stg)}
%     \end{array}
%   \end{smathpar}
The proof follows by structural induction on $c$. Assume that $P
\Leftrightarrow \lambda(\stl,\stg).~\stl=s \conj I(\stg)$ and $Q
\Leftrightarrow \lambda(\stl,\stg).\stl = \F(s,\stg)$. For each type
of $c$, our strategy is to show the premises of the corresponding
RG-rule. The base cases correspond to \C{INSERT}, \C{UPDATE} and \C{DELETE}
statements. In all these cases, the form of $\F$ directly corresponds
to the condition linking the pre-condition $P$ and post-condition $Q$
in the premises of the RG-rules. If $\stable(\R, \F)$, then
$\stabilize{\F} = \F$ and hence the proof directly follows. Otherwise,
$\stabilize{\F} = \lambda (\stg).\existsl(\stg', I(\stg'),
\F( \stg'))$. Since the pre-condition $P(\stl, \stg)$ asserts
$I(\stg)$, any state which obeys $P$ provides the existential $\stg'$
in the above set expression, and hence $\F$ applied on that state
matches the condition linking $P$ and $Q$. The stability of $P$ with
respect to $\R$ follows directly from the stability of $I$.

The proofs for \C{SELECT}, sequencing, and conditionals use the
inductive hypothesis to infer the RG-judgments present in the
premises of their corresponding RG-rules. The interesting case is the
\C{FOREACH} statement, for which we use the loop invariant
$\psi(\stl, \stg) \Leftrightarrow \stl = s \cup (y \bind (\lambda
z.~\F(s, \stg)))$, (where assuming that $c$ is the body of the loop,
$c \elabsto \F$) to prove all the premises of
\rulelabel{RG-Foreach}. Using the same notation as the rule
$\rulelabel{RG-Foreach}$, $y$ refers to the records already processed
in previous iterations of the loop, while $z$ refers to the record
being processed in the current iteration.  At the beginning of the
loop $[\phi/y]\psi(\stl, \stg)$ just reduces to $\stl = s$ which is
implied by the pre-condition $P$. From the inductive hypothesis, we
can infer that each iteration corresponds to the application of
$\F$. Since all iterations are assumed to be independent of each other,
and $z$ is bound to a record in $x$ for each iteration, we conclude
that at the end of every iteration, the loop invariant $[y \cup
  \{z\}/y]\psi$ will be satisfied.

>>>>

We now formally state the correspondence between the inference rules
given above and the RG judgment of \S\ref{sec:reasoning}:
\begin{theorem}
  \label{thm:inference-sound}
  For all $i$,$R$,$I$,$c$,$\Fx$, $\F$, if $\stable(\R,I)$, $\stable(\R, \Fx)$ and $\Fx
  \vdash c \elabsto \F$, then:\\\vspace*{-0.2cm}
  \begin{smathpar}
  \begin{array}{c}
    \R \vdash \hoare{\lambda(\stl,\stg).~\stl=\Fx(\stg) \conj
    I(\stg)}{c}{\lambda(\stl,\stg).~\stl = \Fx(\stg) \cup \F(\stg)}
  \end{array}
  \end{smathpar}
\end{theorem}

\paragraph{{\sc Proof Sketch.}} 
% We prove a slightly stronger statement :
% For all  $i$,$R$,$I$,$c$,$\F$,$s$ if $\stable(\R,I)$ and $c \elabsto \F$, then:
%   \begin{smathpar}
%     \begin{array}{c}
%     \R \vdash \hoare{\lambda(\stl,\stg).~\stl=s \conj
%     I(\stg)}{c}{\lambda(\stl,\stg).\stl = \F(s,\stg)}
%     \end{array}
%   \end{smathpar}
The proof follows by structural induction on $c$. Assume that $P
\Leftrightarrow \lambda(\stl,\stg).~\stl=\Fx(\stg) \conj I(\stg)$ and $Q
\Leftrightarrow \lambda(\stl,\stg).\stl = \Fx(\stg) \cup \F(\stg)$. For each type
of $c$, our strategy is to show the premises of the corresponding
RG-rule. The base cases correspond to \C{INSERT}, \C{UPDATE} and \C{DELETE}
statements. In all these cases, the form of $\F$ directly corresponds
to the condition linking the pre-condition $P$ and post-condition $Q$
in the premises of the RG-rules. If $\stable(\R, \Fx[\F])$, then
$\stabilize{\Fx[\F]} = \F$ and hence the proof directly follows. Otherwise,
$\stabilize{\Fx[\F]} =$ \\%\hspace{2cm}
 $\lambda (\stg).\existsl(\stg', I(\stg'),
\F( \stg'))$. Since the pre-condition $P(\stl, \stg)$ asserts
$I(\stg)$, any state which obeys $P$ provides the existential $\stg'$
in the above set expression, and hence $\F$ applied on that state
matches the condition linking $P$ and $Q$. The stability of $P$ with
respect to $\R$ follows directly from $\stable(\R, I)$ and $\stable(\R, \Fx)$.

The proofs for \C{SELECT}, sequencing, and conditionals use the
inductive hypothesis to infer the RG-judgments present in the
premises of their corresponding RG-rules. The interesting case is the
\C{FOREACH} statement, for which we use the loop invariant
$\psi(\stl, \stg) \Leftrightarrow \stl = \Fx(\stg) \cup (y \bind (\lambda
z.~\F(\stg)))$, (where assuming that $c$ is the body of the loop,
$c \elabsto \F$) to prove all the premises of
\rulelabel{RG-Foreach}. Using the same notation as the rule
$\rulelabel{RG-Foreach}$, $y$ refers to the records already processed
in previous iterations of the loop, while $z$ refers to the record
being processed in the current iteration.  At the beginning of the
loop $[\phi/y]\psi(\stl, \stg)$ just reduces to $\stl = \Fx(\stg)$ which is
implied by the pre-condition $P$. From the inductive hypothesis, we
can infer that each iteration corresponds to the application of
$\F$. Since all iterations are assumed to be independent of each other,
and $z$ is bound to a record in $x$ for each iteration, we conclude
that at the end of every iteration, the loop invariant $[y \cup
  \{z\}/y]\psi$ will be satisfied.
